SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-11-03 2017-Xudong-SMT/ActaInformatica.tex
ACTA Paper: Improve Sect.2

Implementing these semantics raised several challenges:
\begin{itemize}
	\item to obtain a tool that could be applied to pNets representing
	various languages, in particular various action algebras,
	with their specific decision theories;
	\item to separate clearly the 
	algorithm generating the transitions of an open automaton from
	combination of all possible (symbolic) behaviours, and from
	the symbolic reasoning part, specifically here using an SMT
	engine to check the
	satisfiability of the predicates generated by our algorithm;
	\item to build a prototype of the algorithm, and validate the approach on our basic
	case-studies, meanwhile, understand the efficiency of the interaction
	with the SMT solver.
\end{itemize}

@@

\paragraph*{Algebra presentations.}\
In practice, the parameterisation of the pNet model by some specific
action algebra is realised by the definition of a many-sorted algebra
presentation. It will be used to check the
well-formedness of a pNet system, and to define the translation of the pNet
semantics into the SMT engine input language (\cite{BarFT-RR-17}).

@@

\begin{table}[t]\caption{\label{table:BIPalgebra}\leftline{Algebra
			presentation: predefined sorts and operators}}
	\begin{tabular}{p{2.8cm}p{2.8cm}p{5cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
		\\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
		$\texttt{true},\ \texttt{false}$&
		$\land,\ \lor,\ \neg,\ \implies,\ =,\ \ne$
		\\\specialrule{0em}{1pt}{1pt} 
		%               		Action 			&  $\texttt{Synchro},\ \texttt{FUN\_Action\_Bool}$ &
		Action 			&  $\texttt{Synchro},\ \texttt{FUN}$ &
		\\\specialrule{0em}{1pt}{1pt}
		Int 				&
		${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
		$- \texttt{(unary)},\ +,\ -
		\texttt{(binary)},\ \times,\ \div \text{ \etc}$
		%                \\\specialrule{0em}{1pt}{1pt}
		%                \textsl{for any sort} & & $=,\ \ne$
		\\\hline\specialrule{0em}{1pt}{1pt}
		\multicolumn{3}{l}{\sl Extension for the Enable use-case of Section~\ref{section:running-example}}
		\\\hline\specialrule{0em}{1pt}{1pt}
		Action & \texttt{ACT}, \texttt{delta, acc}  &
		\\\hline
	\end{tabular}

@@

@@

@@

@@

\subsection{Example}
\label{section:running-example}

@@

	To give an intuition of the open pNet model and its semantics, we use
	here a simple example based on the ``Enable'' operator of the LOTOS language.
	In the Enable expression ``$P>>Q$'', an \texttt{exit} statement within $P$
	terminates the current process, carrying a value $x$ that
	is captured by the \texttt{accept} statement of $Q$. 	
	We use a simple action algebra,
	containing two constructors $\delta(x)$ and $acc(x)$, for any
	possible data type of the variable $x$, corresponding to the
	statements \texttt{exit} and \texttt{accept}. 
	We need no specific predicate over the action expressions, apart
	from equality of actions.

	In Figure \ref{schema:enable-pnets}, we show two possible encodings of the  
	``Enable'' operator.
        In the first encoding $EnableState$, in the upper part of
	Figure \ref{schema:enable-pnets}, we use a controller $CState$ with two
	states, and simple control actions $l, r, \delta$. The second encoding
	$EnableData$ uses a data-oriented style, with a single state controller $CData$, and a
	state-variable $v$, with values in $\{0,1\}$.

        In this example we use \emph{synchronised
          actions}, to express actions that cannot be further synchronised, similar to the
         silent action $\tau$ of CCS. We denote them as any action expression with the text
        underlined, as e.g. $\underline{\delta(x_2)}$. Such
        synchronised actions do not play any special role for defining
        the semantics of pNets, but as one can expect, will be crucial
        for defining weak equivalences.

	In Figure \ref{schema:enable-composed}, we enrich our example by
	composing two occurences of the $EnableState$ pNet. In Section
        \ref{section:use-case:Enable} we will show how these encodings can
        be used to prove properties of the Enable operator.

@@

\vspace{-5mm}
\begin{figure}[ht]
\centerline{ 
  \includegraphics[width=0.25\linewidth]{ActaXFIG/Enable1-aut}
  \hspace{25mm}
  \includegraphics[width=0.20\linewidth]{ActaXFIG/Enable2-aut}}
  \begin{minipage}{5.7cm}
$  \begin{array}{@{}r@{}l@{}}
    ot_1  = &\openrule{\{ \xrightarrow{a_1}_P \},\\ [\forall y1. a_1\neq\delta(y1)]}
    {A_0 \xrightarrow{a_1} A_0}\\
    ot_2  =& \openrule{
                            \{\xrightarrow{\delta(x_1)}_P,\,
                            \xrightarrow{acc(x_1)}_Q\}  }
    {A_0 \xrightarrow{\underline{\delta(x_1)}} A_1}\\
    ot_3  =& \openrule{
                          \{  \xrightarrow{a_2}_Q \},\\ [\forall y2. a_2\neq acc(y2)] }
    {A_1 \xrightarrow{a_2} A_1}
  \end{array}$
  \end{minipage}
  \hspace{2mm}
  \begin{minipage}{6cm}
$  \begin{array}{@{}r@{}l@{}}
    ot'_1  =& \openrule{
    	\{\xrightarrow{b1}_P\}, ~~ [\forall y1. b1\neq\delta(y1) \!\land\! v=0]}
    {B_0 \xrightarrow{b1} B_0}\\
    ot'_2  =& \openrule{
                            \{\xrightarrow{\delta(y2)}_P,\,
                            \xrightarrow{acc(y2)}_Q \}, \\
                                        v=0, \\
                                        \{v\gets 1\}}
    {B_0 \xrightarrow{\underline{\delta(y2)}} B_1}\\
    ot'_3  =& \openrule{
                            \xrightarrow{b2}_q, \\ [\forall y2. b2\neq\delta(y2) \!\land\! v=1] }
    {B_1 \xrightarrow{b2} B_1}
  \end{array}$
  \end{minipage}
  \caption{The two open automata}
  \label{schema:enable-OAs}
\end{figure}

@@

\begin{figure}[h]
  \begin{minipage}{2cm}
  \includegraphics[width=0.9\linewidth]{ActaXFIG/PQR-automaton}
  \end{minipage}
  \hspace{10mm}
  \begin{minipage}{3.5cm}
  \begin{eqnarray*}
    OT_1 & = \openrule{\{\xrightarrow{a1}_P\}\\ a1\neq\delta(x)}
    {\ostate{00} \xrightarrow{a1} \ostate{00}}\\
    OT_2 & = \openrule{
                            \{\xrightarrow{\delta(x1)}_P, ~~
                            \xrightarrow{acc(x1)}_Q\}
                      }
                      {\ostate{00} \xrightarrow{\underline{\delta(x1)}} \ostate{10}}    \\
    OT_3 & = \openrule{
                        \{\xrightarrow{a2}_Q\},\\a2\neq\delta(x)
                        }
    {\ostate{10} \xrightarrow{a2} \ostate{10}}
    \end{eqnarray*}
  \end{minipage}
  \hspace{10mm}
  \begin{minipage}{4cm}
  \begin{eqnarray*}
    OT_4 & = \openrule{
                      \{\xrightarrow{\delta(x2)}_Q, \qquad 
                       \xrightarrow{acc(x2)}_R\} 
                       }
                      {\ostate{10} \xrightarrow{\underline{\delta(x2)}} \ostate{11}}    \\
    OT_5 & = \openrule{\{\xrightarrow{a3}_R\}
    	}
    {\ostate{11} \xrightarrow{a3} \ostate{11}}
    \end{eqnarray*}
  \end{minipage}
  \caption{Open automaton for the term ``$P>>(Q>>R)$''}
  \label{schema:enable3bis}
\end{figure}

>>>>

Implementing these semantics raised several challenges:
\begin{itemize}
	\item to obtain a tool that could be applied to pNets representing
	various languages, in particular various action algebras,
	with their specific decision theories;
	\item \QIN{to separate clearly the
	transitions of a generated open automaton from the
	combination of all possible (symbolic) behaviours, and from
	the symbolic reasoning part, specifically here, to check the
	satisfiability of the predicates in the transitions generated by our algorithm using an SMT
	engine;}
	\item to build a prototype of the algorithm, and validate the approach on our basic
	case-studies, meanwhile, understand the efficiency of the interaction
	with the SMT solver.
\end{itemize}

@@

\subsection{Algebra Presentations}
In practice, the parameterisation of the pNet model by some specific
action algebra is realised by the definition of a many-sorted algebra
presentation. It will be used to check the
well-formedness of a pNet system, and to define the translation of the pNet
semantics into the SMT engine input language (\cite{BarFT-RR-17}).

@@

\begin{table}[t]\caption{\label{table:BIPalgebra}\leftline{Algebra
			presentation: predefined sorts and operators}}
	\begin{tabular}{p{2.8cm}p{2.8cm}p{5cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
		\\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
		$\texttt{true},\ \texttt{false}$&
		$\land,\ \lor,\ \neg,\ \implies,\ =,\ \ne$
		\\\specialrule{0em}{1pt}{1pt} 
		%               		Action 			&  $\texttt{Synchro},\ \texttt{FUN\_Action\_Bool}$ &
		Action 			&  $\texttt{Synchro},\ \texttt{FUN}$ &
		\\\specialrule{0em}{1pt}{1pt}
		Int 				&
		${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
		$- \texttt{(unary)},\ +,\ -
		\texttt{(binary)},\ \times,\ \div \text{ \etc}$
		%                \\\specialrule{0em}{1pt}{1pt}
		%                \textsl{for any sort} & & $=,\ \ne$
		\\\hline\specialrule{0em}{1pt}{1pt}
		%		\multicolumn{3}{l}{\sl Extension for the Enable use-case of Section~\ref{section:running-example}}
		\multicolumn{3}{l}{\sl Extension for the Enable use-case of Example~\ref{example:EnableOperator}}
		\\\hline\specialrule{0em}{1pt}{1pt}
		%		Action & \texttt{ACT}, \texttt{delta, acc}  &
		Action & \texttt{l}, \texttt{r}, \texttt{$\delta$}, \texttt{acc}  &
		\\\hline
	\end{tabular}

@@

\QIN{
Let us first introduce a notation in LOTOS to give an intuition how we construct the algebra presentation.
}

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\QIN{
Suppose there are two processes $P$, $Q$. In the Enable expression ``$P>>Q$'', an \texttt{exit} statement within $P$ terminates the current process, carrying a value $x$ that is captured by the \texttt{accept} statement of $Q$. 
}

\QIN{
A possible encoding of such ``Enable'' operator, called $EnableState$, uses a controller $CState$ with three simple control actions $l, r, \delta$ in it to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new constructors. Note that the control action $\delta$ is present by the constructor $\delta$ of arity 0 here which is a different application of the constructor.
}

\QIN{
Another possible encoding of such ``Enable'' operator, called $EnableData$, uses a data-oriented style, with a single state controller $CData$, and a state-variable $v$, with values in $\{0,1\}$. We can use the same algebra presentation as $EnableState$ in this case.
}
\end{example}

Next, we enrich our example by composing two occurrences of the $EnableState$. In Section \ref{section:use-case:Enable} we will show how these encodings can be used to prove properties of the Enable operator.
\begin{example}[Composition of Two ``Enable'' Operators]\label{example:ComposeEnableOperators}
\QIN{
Suppose there are three processes $P$, $Q$ and $R$. In the expression ``$P>>(Q>>R)$'', an \texttt{exit} statement within $P$ terminates the current process, carrying a value $x_1$ that is captured by the \texttt{accept} statement of $Q$ in the expression $Q>>R$. And then, similarly, an \texttt{exit} statement within $Q$ terminates the current process, carrying a value $x_2$ that is captured by the \texttt{accept} statement of $R$.
}

\QIN{
We can compose two occurrences of the $EnableState$ to encode such expression. To distinguish the action of two controllers $CState1$ and $CState2$ used in such encoding $EnableStateCompLeft$, we names these control actions $l1, r1, \delta_1$ and $l2, r2, \delta_2$, respectively. The algebra presentation of $EnableStateCompLeft$ can be constructed by expending $\mathit{Constrs}$ with $\{\delta, acc, l1, l2, r1, r2, \delta_1, \delta_2\}$.
}
\end{example}

@@

\begin{example}[PLTS of the Controller in the ``Enable'' Operator]
\QIN{
The PLTS of the controller $CState$ of the encoding $EnableState$ defined in Example \ref{example:EnableOperator} is given as a triple $\langle S,s_0, \to\rangle$ where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$;
	\item[$\bullet$] $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l}0, 0\xrightarrow{\delta}1, 1\xrightarrow{r}1\}$.
\end{itemize}
Add the guard and assignment to the label of the transition, the PLTS of the controller $CData$ is given as a triple $\langle S,s_0, \to\rangle$ where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$;
	\item[$\bullet$] $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l, v=0}0, 0\xrightarrow{\delta,  v=0, \{v\leftarrow1\}}1, 1\xrightarrow{r, v=1}1\}$.
\end{itemize}
}
\end{example}
\QIN{The PLTSs contained in $EnableStateCompLeft$ are similar to the PLTS $CState$.}

@@

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\QIN{
Consider the encoding $EnableState$ defined in Example \ref{example:EnableOperator}, the processes $P$ and $Q$ are holes of the pNet.} In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression with the text underlined, as e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, will be crucial for defining weak equivalences. \QIN{The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	\langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Replace $CState$ with $CData$ the pNet of $EnableData$ can be given with similar $\symb{SV}$.
}
\end{example}

\QIN{The pNet of $EnableStateCompLeft$ presents a bigger tree with three levels containing a pNet as its node.}
\begin{example}[pNet of the Composition of ``Enable'' Operators]\label{example:pNetComposeEnable}
\QIN{
Consider the encoding $EnableStateCompLeft$ defined in Example \ref{example:ComposeEnableOperators}, the processes $P$, $Q$ and $R$ are holes of the pNet. Let $pNet_1$ be the sub-pNet that presents $Q>>R$. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{pNet_1, CState1\}$;
	\item[$\bullet$] $J=\{P\}$;
	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l1\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	\langle \text{-}, a_2, r1\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}.$
\end{itemize}
The sub-pNet $pNet_1$ is a pNet of a single ``Enable'' operator as the pNet in Example \ref{example:pNetEnable}. Actions in $pNet_1$ should be renamed to distinguish themselves from those already used in the pNet of $EnableStateCompLeft$.
}
\end{example}

@@

%\subsection{Example}
%\label{section:running-example}

@@

%	To give an intuition of the open pNet model and its semantics, we use
%	here a simple example based on the ``Enable'' operator of the LOTOS language.
%	In the Enable expression ``$P>>Q$'', an \texttt{exit} statement within $P$
%	terminates the current process, carrying a value $x$ that
%	is captured by the \texttt{accept} statement of $Q$. 	
%	We use a simple action algebra,
%	containing two constructors $\delta(x)$ and $acc(x)$, for any
%	possible data type of the variable $x$, corresponding to the
%	statements \texttt{exit} and \texttt{accept}. 
%	We need no specific predicate over the action expressions, apart
%	from equality of actions.
%        
%	In Figure \ref{schema:enable-pnets}, we show two possible encodings of the  
%	``Enable'' operator.
%        In the first encoding $EnableState$, in the upper part of
%	Figure \ref{schema:enable-pnets}, we use a controller $CState$ with two
%	states, and simple control actions $l, r, \delta$. The second encoding
%	$EnableData$ uses a data-oriented style, with a single state controller $CData$, and a
%	state-variable $v$, with values in $\{0,1\}$.
%
%        In this example we use \emph{synchronised
%          actions}, to express actions that cannot be further synchronised, similar to the
%         silent action $\tau$ of CCS. We denote them as any action expression with the text
%        underlined, as e.g. $\underline{\delta(x_2)}$. Such
%        synchronised actions do not play any special role for defining
%        the semantics of pNets, but as one can expect, will be crucial
%        for defining weak equivalences.
%	
%	Note that synchronisation vectors are defined in a parameterised
%	manner:  the first and third lines represent one vector
%	for each possible action of hole $P$
%	(resp. $Q$). This notation can also use predicates, as in the first
%	case, in which we want the vector to apply to any action of $P$
%	except for $\delta(z)$.
%	
%	In Figure \ref{schema:enable-composed}, we enrich our example by
%	composing two occurences of the $EnableState$ pNet. In Section
%        \ref{section:use-case:Enable} we will show how these encodings can
%        be used to prove properties of the Enable operator.

@@

\vspace{-5mm}
\begin{figure}[ht]
\centerline{ 
  \includegraphics[width=0.25\linewidth]{ActaXFIG/Enable1-aut}
  \hspace{25mm}
  \includegraphics[width=0.20\linewidth]{ActaXFIG/Enable2-aut}}
  \begin{minipage}{5.7cm}
$  \begin{array}{@{}r@{}l@{}}
    ot_1  = &\openrule{\{ \xrightarrow{a_1}_P \},\\ [\forall y1. a_1\neq\delta(y1)]}
    {A_0 \xrightarrow{a_1} A_0}\\
    ot_2  =& \openrule{
                            \{\xrightarrow{\delta(x_1)}_P,\,
                            \xrightarrow{acc(x_1)}_Q\}  }
    {A_0 \xrightarrow{\underline{\delta(x_1)}} A_1}\\
    ot_3  =& \openrule{
                          \{  \xrightarrow{a_2}_Q \},\\ [\forall y2. a_2\neq acc(y2)] }
    {A_1 \xrightarrow{a_2} A_1}
  \end{array}$
  \end{minipage}
  \hspace{2mm}
  \begin{minipage}{6cm}
$  \begin{array}{@{}r@{}l@{}}
    ot'_1  =& \openrule{
    	\{\xrightarrow{b_1}_P\}, ~~ [\forall y1. b_1\neq\delta(y1) \!\land\! v=0]}
    {B_0 \xrightarrow{b_1} B_0}\\
    ot'_2  =& \openrule{
                            \{\xrightarrow{\delta(y_2)}_P,\,
                            \xrightarrow{acc(x_2)}_Q \},~
                                        v=0,~
                                        \{v\gets 1\}}
    {B_0 \xrightarrow{\underline{\delta(x_2)}} B_1}\\
    ot'_3  =& \openrule{
                            \{\xrightarrow{b_2}_Q\}, ~~ [\forall y2. b_2\neq acc(y2) \!\land\! v=1] }
    {B_1 \xrightarrow{b_2} B_1}
  \end{array}$
  \end{minipage}
  \caption{The two open automata}
  \label{schema:enable-OAs}
\end{figure}

@@

\begin{figure}[h]
  \begin{minipage}{2cm}
  \includegraphics[width=0.9\linewidth]{ActaXFIG/PQR-automaton}
  \end{minipage}
  \hspace{10mm}
  \begin{minipage}{3.5cm}
  \begin{eqnarray*}
    OT_1 & = \openrule{\{\xrightarrow{a_1}_P\}\\ a_1\neq\delta(x)}
    {\ostate{00} \xrightarrow{a_1} \ostate{00}}\\
    OT_2 & = \openrule{
                            \{\xrightarrow{\delta(x_1)}_P, ~~
                            \xrightarrow{acc(x_1)}_Q\}
                      }
                      {\ostate{00} \xrightarrow{\underline{\delta(x_1)}} \ostate{10}}    \\
    OT_3 & = \openrule{
                        \{\xrightarrow{a_2}_Q\},\\a_2\neq\delta(x)
                        }
    {\ostate{10} \xrightarrow{a_2} \ostate{10}}
    \end{eqnarray*}
  \end{minipage}
  \hspace{10mm}
  \begin{minipage}{4cm}
  \begin{eqnarray*}
    OT_4 & = \openrule{
                      \{\xrightarrow{\delta(x_2)}_Q, \qquad 
                       \xrightarrow{acc(x_2)}_R\} 
                       }
                      {\ostate{10} \xrightarrow{\underline{\delta(x_2)}} \ostate{11}}    \\
    OT_5 & = \openrule{\{\xrightarrow{a_4}_R\}
    	}
    {\ostate{11} \xrightarrow{a_4} \ostate{11}}
    \end{eqnarray*}
  \end{minipage}
  \caption{Open automaton for the term ``$P>>(Q>>R)$''}
  \label{schema:enable3bis}
\end{figure}
